---
author: 
  "Analysis Team"
date: "`r format(Sys.time(), '%d %B %Y')`"
title    : "High throughput study analysis experiment investigation"


output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    theme: united
    highlight: tango

params:

  inputfile:
    label: "Input files:"
    value: ".Rmd" 
    input: text
  cfgfile:
    label: "cfgfile:"
    value: "config.R"
    input: text    
 

---

```{=html}

<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>
```
<link rel="stylesheet" type="text/css" href="http://cdn.datatables.net/1.10.5/css/jquery.dataTables.min.css">

```{=html}
<script type="text/javascript">
  $(document).ready(function() {
    $("table").DataTable();
  } );
</script>
```

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
    
knitr::opts_chunk$set(warning = FALSE, echo = FALSE)
library(kableExtra)
library(plotly)
library(arsenal)
library(boot)

library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(tidyverse)
library(threejs)
library(pROC)
library(mixtools)

```

## Setting of this report

```{r}
source(params$cfgfile, echo = TRUE)
```

## Aim

Apply Mixture model methods on functional scores. `r RATIO_COMP`.


```{r readin, include=TRUE, warning=FALSE, message=FALSE}
## read in data
load(params$inputfile)

```

## Analysis Plan:

1. Start with the raw counts

2. Get normalized freq and ratio

3. Position adjusted ratio if user defined 

4. Normalization within each exon and across exons

A. Within each exon:

Positional adjusted T2_T0 or T2_T1 ratios from within each replicate were linearly scaled such that the median controls (eg. synonymous and nonsense SNVs) within the replicate were set to the median controls (eg. synonymous and nonsense SNV) values averaged across replicate  experiments.

B. Normalization across exons.

The ensuing SNV scores for each replicate were then normalized across exons in  the same way by again using median controls (eg. synonymous and nonsense SNVs).

C. Functional score

Get the average functional score cross replicates.

5. Apply Mixture Model on Functional score

6. ROC


## Data

In total, there are `r length(unique(norm_ratios$Exon))` Exons in this analysis. 

```{r}

table( norm_ratios$Exon, norm_ratios$Rep)
```

 
Since we have replicates in each Exon, the data plotted below were doubled/tripled or multipled in number. 


## 1. Position adjustment if applicable

Only showing the first Exon before and after. Please refer to the pdf file for all Exons.

```{r, fig.width=10, fig.height=10, warning=FALSE, message=FALSE}
exons <- unique(norm_ratios$Exon)
exons <-  exons[order(as.numeric(sub("E", "", exons)))]
if( LOWESS_ADJ == TRUE) {## there is adjustment
  cbbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

  ratio1 <- paste0("ratio_", RATIO_COMP)
  ratio1.1 <-  paste0(RATIO_COMP, "_loess")
  ratio2 <- paste0("ratio_", RATIO_COMP, "_adjusted")
  ei <- exons[1]
  dti <- norm_ratios %>% filter(Exon == ei)
  if(RATIO_COMP == "T2_T0"){ ## three plots, T1_T0 with lowess line, T2_T0 raw and adjusted
    
    nrep <- unique(dti$Rep)
    nrep <- max(nrep)
    
    ylims <- range(c(dti[, ratio1], dti[, ratio2]), na.rm = T)
    ## for each replicates
    par(mfrow = c(3, 1))
    ## ratio T1_T0 with lowess
    ratio0 <- paste0("ratio_", "T1_T0")
    ratio0.1 <-  paste0("T1_T0", "_loess")
    
    plot(unlist(dti[, ratio0]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio0, col = mycols[dti$Rep], ylim = ylims)
    if(length(grep(ratio0.1, colnames(dti)))>0){
      for(i in 1:nrep){
        lines(unlist(dti[dti$Rep == i, ratio0.1]), x=dti$POS[dti$Rep == i], col = mycols[i])
      }
    }
    
    plot(unlist(dti[, ratio1]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio1, col = mycols[dti$Rep], ylim = ylims)
    # ratio adjusted
    plot(unlist(dti[, ratio2]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio2, col = mycols[dti$Rep], ylim = ylims)
    
  }else{ ## T2_T1 only show two plots
    
    nrep <- unique(norm_ratios$Rep[which(norm_ratios$Exon == ei)])
    nrep <- max(nrep)
    ylims <- range(c(dti[, ratio1], dti[, ratio2]), na.rm = T)
    ## for each replicates
    par(mfrow = c(2, 1))
    ## ratio T2_T1 with loess
    plot(unlist(dti[, ratio1]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio1, col = mycols[dti$Rep], ylim = ylims)
    if(length(grep(ratio1.1, colnames(dti)))>0){
      for(i in 1:nrep){
        lines(unlist(dti[dti$Rep == i, ratio1.1]), x=dti$POS[dti$Rep == i], col = mycols[i])
      }
    }
    ## ratio adjusted
    plot(unlist(dti[, ratio2]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio2, col = mycols[dti$Rep], ylim = ylims)
    
  }
}

```

```{r, message=FALSE}
if( LOWESS_ADJ == TRUE) {## there is adjustment
  ## for each exon put them in pdf. 
  pdf(paste0(OUTDIR, '/', RATIO_COMP,ifelse(LOWESS_ADJ, "_adjusted", ""), '_lowess.pdf'), height = 8, width = 11)
  
  if(RATIO_COMP == "T2_T0"){ ## three plots, T1_T0 with lowess line, T2_T0 raw and adjusted
    for (ei in exons){
      dti <- norm_ratios %>% filter(Exon == ei)
      nrep <- unique(dti$Rep)
      nrep <- max(nrep)
      ylims <- range(c(dti[, ratio1], dti[, ratio2]), na.rm = T)
      ## for each replicates
      par(mfrow = c(3, 1))
      ## ratio T1_T0 with lowess
      ratio0 <- paste0("ratio_", "T1_T0")
      ratio0.1 <-  paste0("T1_T0", "_loess")
  
      plot(unlist(dti[, ratio0]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio0, col = mycols[dti$Rep], ylim = ylims)
      if(length(grep(ratio0.1, colnames(dti)))>0){
        for(i in 1:nrep){
          lines(unlist(dti[dti$Rep == i, ratio0.1]), x=dti$POS[dti$Rep == i], col = mycols[i])
        }
      }
      
      plot(unlist(dti[, ratio1]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio1, col = mycols[dti$Rep], ylim = ylims)
      # ratio adjusted
      plot(unlist(dti[, ratio2]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio2, col = mycols[dti$Rep], ylim = ylims)
    }
  }else{ ## T2_T1 only show two plots
    for (ei in exons){
      dti <- norm_ratios %>% filter(Exon == ei)
      nrep <- unique(dti$Rep)
      nrep <- max(nrep)
      ylims <- range(c(dti[, ratio1], dti[, ratio2]), na.rm = T)
      ## for each replicates
      par(mfrow = c(2, 1))
      ## ratio T2_T1 with loess
      plot(unlist(dti[, ratio1]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio1, col = mycols[dti$Rep], ylim = ylims)
      if(length(grep(ratio1.1, colnames(dti)))>0){
        for(i in 1:nrep){
          lines(unlist(dti[dti$Rep == i, ratio1.1]), x=dti$POS[dti$Rep == i], col = mycols[i])
        }
      }
      ## ratio adjusted
      plot(unlist(dti[, ratio2]), x=unlist(dti$POS), type="p", pch = 20, cex = 0.2, main=ei, xlab="Position", ylab= ratio2, col = mycols[dti$Rep], ylim = ylims)
    }
  }
  
  
  dev.off()
  
}
```

## 2. Normalized Results for Ratio


### Step1 normalization within exon


```{r, include=TRUE, fig.width=4*min(length(exons), 3), fig.height=4*ceiling(length(exons)/3)}
score0 <- paste0("ratio_", RATIO_COMP, ifelse(LOWESS_ADJ, "_adjusted", ""))
score1 <- paste0(RATIO_COMP,ifelse(LOWESS_ADJ, "_adjusted", ""), "_norm_wie")
score2 <- paste0(RATIO_COMP,ifelse(LOWESS_ADJ, "_adjusted", ""), "_norm_xe")


etype8 <- unique(norm_ratios$EventType)
if(length(etype8) >= 8){
  mycol8 <-  c(1, 5, 6, 2, 3, 4, 7:length(etype8))
}else{
  mycol8 <-  c(1, 5, 3, 4, 2, 6)

}
names(mycol8) <- etype8
mycol2.1 <- mycol8[etype8]

nrows <- max(ceiling(length(exons)/3), 1)

par(mfrow=c(nrows, 3))

for(ei in exons){
  dt2plot <- norm_ratios %>% filter(Exon == ei) %>% 
    mutate(R0 = !!sym(score0),
           R1 = !!sym(score1),
           R2 = !!sym(score2))
  temp <- mycol2.1[match(dt2plot$EventType, etype8)]
  plot(dt2plot$R0, dt2plot$R1, pch = 20,
       main = paste0(ei, " (N =", max(dt2plot$Rep), " , P =",  length(unique(dt2plot$POS)), ")"), col = temp,
       xlab = "before", ylab = 'after')
  abline(a = 0, b = 1, col = "gray", lty = 2)
}
plot(0, 0, type = "n", xlab = '', ylab = '', axes = F)
legend(x = "bottom", 
       legend = etype8,
       pch =20,
       col =mycol2.1,
       xpd = TRUE, 
       cex = 1, ncol = 1) # Horizontal legend
```

### Step2 normalization across exons.


```{r, include=TRUE, fig.width=4*min(length(exons)+1, 3), fig.height=4*ceiling(length(exons)/3)}

par(mfrow=c(nrows, 3))

for(ei in exons){
  dt2plot <- norm_ratios %>% filter(Exon == ei) %>% 
    mutate(
           R1 = !!sym(score1),
           R2 = !!sym(score2))
  temp <- mycol2.1[match(dt2plot$EventType, etype8)]
  plot(dt2plot$R1, dt2plot$R2, pch = 20,
       main = paste0(ei, " (N =", max(dt2plot$Rep), " , P =",  length(unique(dt2plot$POS)), ")"), col = temp,
       xlab = "before", ylab = 'after')
  abline(a = 0, b = 1, col = "gray", lty = 2)
}
plot(0, 0, type = "n", xlab = '', ylab = '', axes = F)
legend(x = "bottom", 
       legend = etype8,
       pch =20,
       col =mycol2.1,
       xpd = TRUE, 
       cex = 1, ncol = 1) # Horizontal legend

```

### Put together 

Ratio -> within exon normalized Ratio -> cross exon normalized Ratio


```{r, include=TRUE, fig.width=10, fig.height=15}
dt2plot <- norm_ratios %>% filter(!is.na( !!sym(score0)))
par(mfrow=c(4,1))
myylims <- range(unlist(dt2plot[, score0]), na.rm = TRUE)
plot(dt2plot$POS, unlist(dt2plot[, score0]), pch = 20, col = mycol8[dt2plot$EventType], main = "Raw", 
     xlab = "genomic position", ylab = 'ratio', ylim = myylims)
abline(h = 0, col = "gray", lty = 2)

myylims <- range(unlist(dt2plot[, score1]), na.rm = TRUE)
plot(dt2plot$POS, unlist(dt2plot[, score1]), pch = 20, col = mycol8[dt2plot$EventType], main = "Normalization within Exon", 
     xlab = "genomic position", ylab = 'wie normalized ratio', ylim = myylims)
abline(h = 0, col = "gray", lty = 2)

myylims <- range(unlist(dt2plot[, score2]), na.rm = TRUE)
plot(dt2plot$POS, unlist(dt2plot[, score2]), pch = 20, col = mycol8[dt2plot$EventType], main = "Normalization across Exons", 
     xlab = "genomic position", ylab = 'xe normalized ratio', ylim = myylims)
abline(h = 0, col = "gray", lty = 2)
plot(0, 0, type = "n", xlab = '', ylab = '', axes = F)
legend(x = "bottom", 
       legend = etype8,
       pch =20,
       col =mycol2.1,
       xpd = TRUE, 
       cex = 1, ncol = 3) # Horizontal legend

```


## 3. Functional Score distribution

### Normality check 

Plot shows controls only, including splicing region variants if there is any

```{r}

dt2plot <- FS_data %>% filter(EventType %in% c(POS_GROUP, NEG_GROUP)) 
p <- dt2plot %>%
  ggplot( aes(x=functional_score, fill=EventType)) +
  geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity', binwidth = 0.20) +
  scale_fill_manual(values=c( "red", "#69b3a2")) +  ggtitle ( paste0(RATIO_COMP," Functional Score"))
ggplotly(p)

```


Shapiro-Wilk Normality test


`r POS_GROUP` positive control group


```{r}

pos <- dt2plot$functional_score[which(dt2plot$EventType %in% POS_GROUP)]
if(length(pos) > 2 & length(pos) < 5000) {
  shapiro.test(pos[sample(1:length(pos), min(4500, length(pos)))])
}else{
  cat("! sample size must be between 3 and 5000\n")
}
```

`r NEG_GROUP` negative control group

```{r}

neg <- dt2plot$functional_score[which(dt2plot$EventType %in% NEG_GROUP)]

if(length(neg) > 2 & length(neg) < 5000) {
  shapiro.test(neg[sample(1:length(neg), min(4500, length(neg)))])
}else{
  cat("! sample size must be between 3 and 5000\n")
}
```


### QQ plots 


```{r, fig.width=9, fig.height=5}

par(mfrow=c(1,2))
qqnorm(neg, main=paste0(RATIO_COMP, "_", NEG_GROUP))
qqline(neg, col = "steelblue", lwd = 2)
qqnorm(pos, main=paste0(RATIO_COMP, "_", POS_GROUP))
qqline(pos, col = "steelblue", lwd = 2)


```

### Group distribution

Include all variants. by `r TYPE_COLUMN`

```{r, fig.width=9, fig.height=5}
dt2plot <- FS_data %>% filter(!is.na(functional_score)) 

p <- dt2plot %>%
  ggplot( aes(x=functional_score, fill=EventType)) +
  geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity', binwidth = 0.15) +  ggtitle ( paste0(RATIO_COMP," functional score"))
ggplotly(p)

```

```{r, fig.width=9, fig.height=5}
ggplot(dt2plot, aes(x=EventType, y=functional_score)) + 
  geom_boxplot(outlier.colour="red", outlier.shape=8,
                outlier.size=4) +  ggtitle ( paste0(RATIO_COMP," functional score")) + geom_jitter(shape=16, position=position_jitter(0.2))+
  theme(axis.text.x = element_text(angle = 30, hjust = 1))


```

### Exon FS distribution 

```{r, fig.width=9, fig.height=5}
dt2plot <- FS_data %>% filter(!is.na(functional_score)) %>% filter(EventType %in% c(NEG_GROUP, POS_GROUP)) 
p <- dt2plot %>% 
  ggplot( aes(x=functional_score, y = Exon, fill=EventType, text = paste0(POS, "_", REF, "_", ALT))) +
  geom_point( color="#e9ecef", alpha=0.6, position = 'identity', binwidth = 0.1) +
  scale_fill_manual(values=c( "red", "#69b3a2")) +  ggtitle (paste0(RATIO_COMP," functional score"))
ggplotly(p)

```

## 5. Classification perf 


### 5.1 ROC 

on all variants

```{r ROC1.av, include=TRUE, fig.width=5, fig.height=5, message=FALSE, warning=FALSE}
dt2plot <- FS_data %>% filter(!is.na(functional_score)) %>% filter(EventType %in% c(NEG_GROUP, POS_GROUP)) %>% mutate(labels = ifelse(EventType %in% c(NEG_GROUP), "control", "case"))
ROC0.av <- calculate_auc_balance (predictions=dt2plot$functional_score, 
                       labels = dt2plot$labels, 
                       plot = TRUE, plot_title = "ROC Curve") 
ROC0.av$plot
```

Classification cutoff `r round(ROC0.av$balance_threshold, 3)`, sensitivity `r round(ROC0.av$sensitivity, 2)`, specificity `r round(ROC0.av$specificity, 2)`.


### 5.2 Mixture Model 

Findlay paper "fitting a two-component Gaussian mixture model in which the parameters of the ‘non-functional’ distribution were based on all nonsense SNVs and the ‘functional’ distribution based on synonymous SNVs not depleted in RNA (Extended Data Fig. 7). We then used this model to estimate the probability of each SNV’s score being drawn from the non-functional distribution (Pnf). SNVs with Pnf < 0.01 were categorized as functional (72.5%); SNVs with Pnf > 0.99 were categorized as non-functional (21.1%); and SNVs with 0.01 < Pnf < 0.99 (6.4%) were categorized as intermediate."



```{r MixModel, include=TRUE, fig.width=9, fig.height=5, message=FALSE, warning=FALSE}

mymm$plot1
mymm$plot2
```


`r paste0("There is",  ifelse(mymm$error.occured, " AN ", " NO "), "Error Ocurred")`.

`r mymm$f99` was the functional score corresponding to the probability of non functional at 99% and `r mymm$f1` was the functional score corresponding to the probability of non functional at 1%.

`r mymm$f98` was the functional score corresponding to the probability of non functional at 98% and `r mymm$f2` was the functional score corresponding to the probability of non functional at 2%.

`r mymm$f95` was the functional score corresponding to the probability of non functional at 95% and `r mymm$f5` was the functional score corresponding to the probability of non functional at 5%.

`r mymm$f90` was the functional score corresponding to the probability of non functional at 90% and `r mymm$f10` was the functional score corresponding to the probability of non functional at 10%.

`r mymm$f80` was the functional score corresponding to the probability of non functional at 98% and `r mymm$f20` was the functional score corresponding to the probability of non functional at 20%.




